Retrofitting Effect Handlers onto OCaml

### 2021-03-23 retro-concurrency.tex
updates

%\vspace{-1em}
\subsection{Asynchronous I/O}
\label{sec:aio}

Effect handlers provide a structured form of delimited
continuations~\cite{Danvy90} that enables non-local control-flow mechanisms
such as resumable exceptions, lightweight threads, coroutines, generators and
asynchronous I/O to be composably expressed. One of the primary motivations to
extend OCaml with effect handlers is to support concurrent programming. To
illustrate this, let us implement a simple lightweight thread library with
support for forking new threads and yielding control to other threads. We will
then extend this scheduler with support for synchronously reading from channels
and subsequently make it asynchronous without changing the client code for
asynchrony.

Many programming languages, including OCaml, require non-local changes to source
code in order to support asynchronous I/O, often leading to a dichotomy between
synchronous and asynchronous code~\cite{colour}.
Since OCaml only provides preemptive system threads as support for concurrency,
users usually use cooperative I/O libraries such as Lwt~\cite{lwt} and
Async~\cite{async}, where all asynchronous code must
use a monadic style. Languages such as GHC Haskell and Go provide lightweight
threads, which avoids the dichotomy between synchronous and asynchronous code.
However, these languages bake-in the lightweight thread implementation into the
runtime system. With effect handlers, asynchronous I/O can be implemented
directly in OCaml as a library without imposing a monadic form on users of it.

%\vspace{-1.5mm}
\subsection{Retrofitting effect handlers}

There are many research languages and libraries built around effect
handlers~\cite{Leijen14,Hillerstrom20,Pyro,Frank,Eff}. Unlike these efforts,
our goal is to retrofit effect handlers onto the OCaml programming language,
which has been in continuous use for the past 25 years in large codebases
including verification tools~\cite{everest,Coq}, mission critical software
systems~\cite{astree} and latency sensitive networked
applications~\cite{Madhavapeddy13}. OCaml is particularly favoured for its
competitive yet predictable performance, with a fast foreign-function interface
(FFI). It has excellent compatibility with program analysis tools such as
debuggers and profilers that utilise DWARF stack unwind tables~\cite{DWARF} to
obtain a backtrace. Our primary motivation for adding effect handlers to OCaml
is to support: {\em (i)} highly concurrent applications such as web servers
using a natural \emph{direct-style} as opposed to using callbacks; and {\em
(ii)} efficient nested parallel programs in the vein of Cilk~\cite{Cilk},
OpenMP~\cite{OpenMP}, X10~\cite{Charles05}, Intel TBB~\cite{IntelTBB}, and
variants of Parallel ML~\cite{MaPLe, Fluet10, Sivaramakrishnan14}.

OCaml currently does not support any non-local control flow mechanisms other
than exceptions. This makes it particularly challenging to implement the
delimited continuations necessary for effect handlers without sacrificing the
desirable properties of OCaml. One standard way of implementing continuations
is to use continuation-passing style (CPS) in the compiler's intermediate
representation (IR)~\cite{Leijen14}. OCaml does not use a CPS IR, and changing
the compiler to utilise a CPS IR would be an enormous undertaking that would
affect the performance profile of existing OCaml applications due to the
increased memory allocations as the continuation closures get allocated on the
heap~\cite{Farvardin20}. Moreover, with CPS, stack unwinding is meaningless and
hence, we would lose compatibility with tools that inspect the program stack.
Hence, we choose not to use CPS translation and represent the continuations as
call stacks.

The search for an expressive effect system that guarantees that all the effects
performed in the program are handled (\emph{effect safety}) in the presence of
advanced features such as polymorphism, modularity and generativity is an
active area of research~\cite{Leijen14, Biernacki19, Biernacki20,
Hillerstrom20}. We do not focus on this question in this paper, and our
implementation of effect handlers in OCaml does not guarantee effect safety.
For example, running |perform (Read stdin)| without an enclosing handler raises
|Unhandled| exception at the point of |perform|. We leave the question of
effect safety for future work.

>>>>

Effect handlers provide a structured form of delimited
continuations~\cite{Danvy90} that enables non-local control-flow mechanisms
such as resumable exceptions, lightweight threads, coroutines, generators and
asynchronous I/O to be composably expressed. One of the primary motivations to
extend OCaml with effect handlers is to natively support composable
asynchronous I/O. Many programming languages, including OCaml, require
non-local changes to source code in order to support asynchronous I/O, often
leading to a dichotomy between synchronous and asynchronous code~\cite{colour}.
In the absence of native support for concurrent programming, OCaml developers
use monadic I/O libraries such as Lwt~\cite{lwt} and Async~\cite{async}, where
asynchronous functions are represented as monadic computations. In these
libraries, while asynchronous functions can call synchronous functions
directly, the converse is not true. In particular, any function that calls an
asynchronous function will also have to be marked as asynchronous. Languages
such as GHC Haskell and Go provide lightweight threads, which avoids the
dichotomy between synchronous and asynchronous code. However, these languages
bake-in the lightweight thread implementation into the runtime system. With
effect handlers, asynchronous I/O can be implemented directly in OCaml as a
library without imposing a monadic form on the users.

There are many research languages and libraries built around effect
handlers~\cite{Leijen14,Hillerstrom20,Pyro,Frank,Eff}. Unlike these efforts,
our goal is to retrofit effect handlers onto the OCaml programming language,
which has been in continuous use for the past 25 years in large codebases
including verification tools~\cite{everest,Coq}, mission critical software
systems~\cite{astree} and latency sensitive networked
applications~\cite{Madhavapeddy13}. OCaml is particularly favoured for its
competitive yet predictable performance, with a fast foreign-function interface
(FFI). It has excellent compatibility with program analysis tools such as
debuggers and profilers that utilise DWARF stack unwind tables~\cite{DWARF} to
obtain a backtrace.

OCaml currently does not support any non-local control flow mechanisms other
than exceptions. This makes it particularly challenging to implement the
delimited continuations necessary for effect handlers without sacrificing the
desirable properties of OCaml. One standard way of implementing continuations
is to use continuation-passing style (CPS) in the compiler's intermediate
representation (IR)~\cite{Leijen14}. OCaml does not use a CPS IR, and changing
the compiler to utilise a CPS IR would be an enormous undertaking that would
affect the performance profile of existing OCaml applications due to the
increased memory allocations as the continuation closures get allocated on the
heap~\cite{Farvardin20}. Moreover, with CPS, stack unwinding is meaningless,
and hence, we would lose compatibility with tools that inspect the program
stack. Hence, we choose not to use CPS translation and represent the
continuations as call stacks.

The search for an expressive effect system that guarantees that all the effects
performed in the program are handled (\emph{effect safety}) in the presence of
advanced features such as polymorphism, modularity and generativity is an
active area of research~\cite{Leijen14, Biernacki19, Biernacki20,
Hillerstrom20}. We do not focus on this question in this paper, and our
implementation of effect handlers in OCaml does not guarantee effect safety. We
leave the question of effect safety for future work.
